---
title: "Problem Set 1"
author: "Nico Hawley-Weld"
format: markdown_github
date: 2024-01-27
---

# Problem 1
Define variables $a=1$, $b=-1$, $c=-2$ and print out the solutions to $f(x) = ax^2 + bx + c = 0$. Do not report complex solutions, only real numbers. Avoid using the variable name `c` as it is a reserved function in R. Show the code and the answer.

```{r}
# Define function to solve for a, b, and c2
solve_quadratic <- function(a, b, c2) {
    # Calculate discriminant
    discriminant <- b^2 - 4*a*c2

    # Check if the discriminant is non-negative
    if (discriminant >= 0) {
        # Calculate the two real solutions
        x1 <- (-b + sqrt(discriminant)) / (2*a)
        x2 <- (-b - sqrt(discriminant)) / (2*a)

        # Print the solutions
        cat("The solutions are:", x1, "and", x2, "\n")
    } else {
        cat("There are no real solutions.\n")
    }
}

# Solve for a=1, b=-1, c=-2
solve_quadratic(1,-1,-2)
```

# Problem 2
Show a graph of $f(x)$ versus $x$ for $x \in (-3,3)$. Do not show the code, only the graph.

```{r echo=FALSE}
# Define plotting function
plot_quadratic <- function(a, b, c2, xmin, xmax) {

  # Define quadratic function
  f <- function(x) {
  a*x^2 + b*x + c2
  }

  # Plot the function
  plot(f, xmin, xmax, main = "Graph of f(x) = ax^2 + bx + c2", xlab = "x", ylab = "f(x)")  
  
  # Add a grid for better visualization
  grid()
}

# Plot a=1, b=-1, c=-2, xmin=-3, xmax=3
plot_quadratic(1,-1,-2,-3,3)
```

# Problem 3
Generate a PDF report.

# Problem 4
Erase the PDF report and reproduce it but this time using $a=1$, $b=2$, $c=5$.

```{r echo=FALSE}
# Solve for and plot a=1, b=2, c=5
solve_quadratic(1,2,5)
plot_quadratic(1,2,5,-3,3)
```

# Problem 5
Erase the PDF report and reproduce it but this time using $a=1$, $b=3$, $c=2$. Change the range of `x` to range that clearly shows the roots.

```{r echo=FALSE}
# Solve for and plot a=1, b=3, c=2
solve_quadratic(1,3,2)
plot_quadratic(1,3,2,-3,0)
```

# Problem 6
Create a markdown page with the results for this last set of values, but this time showing the code.

```{r}
# Solve for and plot a=1, b=3, c=2
solve_quadratic(1,3,2)
plot_quadratic(1,3,2,-3,0)
```

# Problem 7
Submit the markdown page, including all necessary auxiliary files, and quarto file to a GitHub repo. Make sure the markdown document renders.